Modelling last glacial cycle ice dynamics in the Alps

### 2018-08-16 articles/cycle/response-editor1.tex

% response-editor.tex
% ----------------------------------------------------------------------
\input{response-header}

\title{Authors' response to the Editor}
\author{J.~Seguinot et al.}
\date{February 15, 2018}

\begin{document}
\maketitle
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\referee}[2][]{\bigskip\noindent\textcolor{darkblue}{\textbf{#1:} #2}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}

Dear Andreas Vieli,

Thank you very much for your feedback and support for our manuscript.

    \referee[p.~1, l.~6]{%
        to be correct it should probably say: '...resolution of 1 to 2 km'.}

    We rephrased to ``using horizontal resolutions of 2 and 1\,km and up to
    576~processors''.

    \referee[p.~2, l.~12]{%
        '...indicating SEVERAL tens of ...'. And change 800 000 years to 800 ka
        (and delete the bracket (800ka) on next line.}

    Done.

    \referee[p.~5, Fig.~1c caption]{%
        maybe say that this is at the 1km resolution used in the model.}

    Good idea, we've added this precision.

    \referee[p.~11, l.~6]{%
        '...by a east-west...'}

    Thanks, we've corrected.

    \referee[p.~15]{%
        I am just wondering and maybe this is to be considered in a later stage
        after a first round of reviews: in the section 4.2 i wonder if you
        should not briefly consider the discrepancy of the erratics from the
        southern Valais that reached the solothurner lobe... (work of Jouvet et
        al). it seems with your modelling you do not get Valais ice to flow to
        Solothurn (only aare valley)}

    Although not entirely obvious from the figures this is a correct
    observation. We've added a sentence about this.

    \referee[p.~18, Fig.~6 caption l.~4]{%
        should it not say 50m bands (instead of 500m bands)?}

    That is right, we've corrected.

    \referee[p.~18, l.~5--7]{%
        should you not mention the work by Becker here (his Phd work), as he
        also did preliminary sensitivity studies exploring a range of factors
        parameters (including sliding/rheology) for a part of the alpine ice
        sheet. Or am I wrong here?}

    We've added a sentence about that.

\end{document}

### 2018-08-16 articles/cycle/response-editor3.tex

% response-editor.tex
% ----------------------------------------------------------------------
\input{response-header}

\title{Authors' response to the Editor}
\author{J. Seguinot, on behalf of all authors.}
%\date{}

\begin{document}
\thispagestyle{empty}
\maketitle
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

    Dear Andreas Vieli,

    \newcommand{\referee}[1]{\bigskip\noindent\textcolor{journalname}{#1}}

    \referee{\textbf{p.~1, l.~14-17:}
        somewhat odd phrasing: `...cause ...to be modelled...', maybe
        `...produces in the model...' or similar would be better.}

    \referee{\textbf{p.~3, l.~4:}
        `it still remains incompletely known' is awkward phrasing, maybe better
        say: `...knowledge remains incomplete...'}

    \referee{\textbf{p.~3, l.~7-8:}
        the added sentence starting with `Additional...' seems slightly out of
        place here (or the flow is someaht disrupted). Maybe better move it to
        end of nextz paragraph (p.~3, l.~14).}

    \referee{\textbf{p.~3, l.~9:}
        starting a paragraph with a `But...' is awkward, maybe change back to
        `Here we...' in particular when you move the sentence before.}

    \referee{\textbf{p.~3, l.~32:}
        would `different' instead of `distinct' not be more appropriate.}

    \referee{\textbf{p.~4, l.~8:}
        `In the SIA, topographic roughness using a bed smoother range of 5 km
        (Schoof, 2003).' I really do not understand this sentence (what is
        message), nor its connection to the sentence before.}

    \referee{\textbf{p.~4, l.~24:}
        `...without horizontal transport...'?}

    \referee{\textbf{p.~4, l.~25:}
        maybe simplify to `...is assumed to drain instantaneously and is
        removed....'}

    \referee{\textbf{p.~7, l.~3:}
        `above average' of what? What is reference?}

    \referee{\textbf{p.~14, fig.~4:}
        are the surface contours really in 200m intervals?, that would make the
        ice surface only a bit over 1200m high. Or do you leave some out?  You
        should add some labels to the surface contour lines so one can judge
        the surface elevation better.}

    \referee{\textbf{p.~13, l.~30 -- p.~14, l.~3:}
        but jouvet (TC) claims (based on tracking erratics) that north south
        contrast in in precip is needed. Is this still consistent with this
        work? Maybe should be included. This is mentioned on next page but
        would be relevant here as well.}

    \referee{\textbf{p.~14, fig.~4 caption:}
        for `(b)' it should say what the vertical black line (time of max
        extent in (a) at 24.57 ka???) refers to.}

    \referee{\textbf{p.~15, l.~31:}
        `simulation' (not sumulation)}

    \referee{\textbf{p.~16, l.~12:}
        `Fig. 5' has nothing to do with `reviews', what do you mean with `Fig.
        5 for reviews'???}

    \referee{\textbf{p.~19, l.~15-19:}
        in this context of sensitivity to basal sliding parameter (and lacking
        exploration) maybe some reference to the Phd-thesis of Becker which
        includes at least some preliminary (mostly steady state) sliding
        investigation could be made.}

    \referee{\textbf{p.~21, l.~19:}
        `last major Alpine reliefs' ...? Do you mean the `last foothills' of the
        alps? Clarify.}

    \referee{\textbf{p.~23, l.~1:}
        is it not rather `glacier physics' as the till is not ice (or ice and
        glacier physics).}

    We thank you again very much for your editorial work on our manuscript.

\end{document}

### 2018-08-16 articles/cycle/response-editor3.tex

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\noindent\textcolor{journalname}{#1}}

    Dear Andreas Vieli,

    Thank you very much for supporting our manuscript for publication in
    \emph{The Cryosphere}. We have adressed your comments and performed a few
    last changes on the manuscript as detailed below.

% ----------------------------------------------------------------------

\sechead{Editor comments}

@@

    \referee{\textbf{p.~3, l.~32:}
        would `different' instead of `distinct' not be more appropriate.}

    We replaced with ``different''.

    \referee{\textbf{p.~4, l.~8:}
        `In the SIA, topographic roughness using a bed smoother range of 5 km
        (Schoof, 2003).' I really do not understand this sentence (what is
        message), nor its connection to the sentence before.}

    Sorry, two words were missing here: ``topographic roughness is parametrized
    using a bed smoother range of 5\,km.'' We have moved this sentence directly
    after the previous mention of the SIA.

@@

    \referee{\textbf{p.~4, l.~25:}
        maybe simplify to `...is assumed to drain instantaneously and is
        removed....'}

    Done.

    \referee{\textbf{p.~7, l.~3:}
        `above average' of what? What is reference?}

    We have changed to ``above annual lapse rates measured in the Alps''. The
    reference describes lapse rate measurements.

@@

    \referee{\textbf{p.~14, fig.~4:}
        are the surface contours really in 200m intervals?, that would make the
        ice surface only a bit over 1200m high. Or do you leave some out?  You
        should add some labels to the surface contour lines so one can judge
        the surface elevation better.}

    An incorrect version of the figure missing 200m contours was uploaded. This
    has been corrected. Thank you very much for noticing.

@@

    \referee{\textbf{p.~13, l.~30 -- p.~14, l.~3:}
        but jouvet (TC) claims (based on tracking erratics) that north south
        contrast in in precip is needed. Is this still consistent with this
        work? Maybe should be included. This is mentioned on next page but
        would be relevant here as well.}

    The reference was added. However, to avoid redundancy with the next section
    on flow patterns, we would prefer to limit the discussion in this section
    to modelled ice marginal positions.

@@

    \referee{\textbf{p.~14, fig.~4 caption:}
        for `(b)' it should say what the vertical black line (time of max
        extent in (a) at 24.57 ka???) refers to.}

    We have made explicit that: ``The black vertical line indicates the
    modelled age of maximum ice cover at 24.57\,ka.''

@@

    \referee{\textbf{p.~15, l.~31:}
        `simulation' (not sumulation)}

    Done.

    \referee{\textbf{p.~16, l.~12:}
        `Fig. 5' has nothing to do with `reviews', what do you mean with `Fig.
        5 for reviews'???}

    With ``cf.~Ivy-Ochs, 2015; Wirsig et al., 2016, Fig. 5 for reviews'', we
    meant to refer to Fig.~5 in the paper by Wirsig et al. (2016). We have
    removed ``Fig.~5'' in this particular instance.

@@

    \referee{\textbf{p.~19, l.~15-19:}
        in this context of sensitivity to basal sliding parameter (and lacking
        exploration) maybe some reference to the Phd-thesis of Becker which
        includes at least some preliminary (mostly steady state) sliding
        investigation could be made.}

    Done.

    \referee{\textbf{p.~21, l.~19:}
        `last major Alpine reliefs' ...? Do you mean the `last foothills' of the
        alps? Clarify.}

    We have replaced this formulation with ``the outermost limestone reliefs''.

@@

    \referee{\textbf{p.~23, l.~1:}
        is it not rather `glacier physics' as the till is not ice (or ice and
        glacier physics).}

    Correct. We have changed to ``glacier physics''.

% ----------------------------------------------------------------------

%\sechead{New references}

### 2018-08-16 articles/cycle/alps-cycle.tex
Adress last comments from the editor.

    Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
    glacial cycle (120--0\,ka) in the Alps, using horizontal resolutions of 2
    and 1\,km. Climate forcing is derived using present-day climate data from
    WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
    offsets from multiple palaeo-climate proxies, among which only the EPICA ice
    core record yields glaciation during marine oxygen isotope stages~4
    (69--62\,ka) and~2 (34--18\,ka) spatially and temporally consistent with
    the geological reconstructions, while the other records used result in
    excessive early glacial cycle ice cover and a late Last Glacial Maximum.
    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that Alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences through the
    mountain passes. Modelled maximum ice surface is in average 861\,m
    higher than observed trimline elevations in the upper Rhone Valley, yet
    our simulation predicts little erosion at high elevation due to cold-based
    ice. Finally, despite the uniform climate forcing, differences in glacier
    catchment hypsometry cause the Last Glacial Maximum advance to be modelled
    as a time-transgressive event, with some glaciers reaching their maximum as
    early as 27\,ka, and others as late as 21\,ka.

@@

    Although the glacial history of the European Alps has been studied for
    nearly three hundred years, it still remains incompletely known
      (1) what climate evolution led to the known maximum ice limits,
      (2) to what extent ice flow was controlled by subglacial topography,
      (3) what drove the different responses of the individual lobes,
      (4) how far above the trimline was the ice surface located, and
      (5) how many advances occurred during the last glacial cycle.
    Additional geological research will be needed to complete our knowledge.

    But here, we intend to explore these open questions from a new angle and
    use the Parallel Ice Sheet Model
    \citep[PISM,][]{PISM-authors.2017}, a numerical ice sheet model that
    approximates glacier sliding and deformation (Sect.~\ref{sec:model}), to
    model Alpine glacier dynamics through the last glacial cycle (120--0\,ka),
    a period for which palaeo-temperature proxies are available, albeit non
    regional. In an attempt to analyse the longstanding questions outlined
    above, we test the model sensitivity to multiple palaeo-climate forcing
    (Sect.~\ref{sec:climate}), and then explore the modelled glacier dynamics
    at high resolution for the optimal forcing (Sect.~\ref{sec:results}).

@@

    Each simulation starts from assumed present-day ice thickness and
    equilibrium ice and bedrock temperature at 120\,ka, and runs to the
    present. Our modelling domain of 900 by 600\,km encompasses the
    entire Alpine range (Fig.~\ref{fig:inputs}). The simulations were run on
    two distinct grids, using horizontal resolutions of 2 and 1\,km,
    respectively.
    All parameter values are summarized in Table~\ref{tab:params}.

@@

    Ice sheet dynamics are typically modelled using a~combination of internal
    deformation and basal sliding. PISM is a~shallow ice sheet model, which
    implies that the balance of stresses is approximated based on their
    predominant components. The Shallow Shelf Approximation (SSA) is combined
    with the Shallow Ice Approximation (SIA) by adding velocity solutions of
    the two approximations \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
    Although this heuristic approach is inaccurate in the transition zone
    where both vertical shear and longitudinal stretch are significant,
    it was shown to effectively approximate complex ice sheet flow
    dynamics over mountainous topographies similar to that of the European Alps
    \citep{Golledge.etal.2012, Ziemen.etal.2016}. In the SIA, topographic
    roughness using a bed smoother range of 5\,km \citep{Schoof.2003}.

@@

    A~pseudo-plastic sliding law relates the bed-parallel shear stresses to the
    sliding velocity. The yield stress is modelled using the Mohr--Coulomb
    criterion \citep[Table~\ref{tab:params};][Eq.~18]{Tulaczyk.etal.2000}.
    using a constant basal friction angle corresponding to the average of
    available measurements \citep[p.~268]{Cuffey.Paterson.2010}.
    Effective pressure is related to the ice overburden stress and the modelled
    amount of subglacial water, using a~formula derived from laboratory
    experiments with till extracted from the base of Ice Stream B in West
    Antarctica \citep[Table~\ref{tab:params};][Eqs~23 and
    24]{Tulaczyk.etal.2000, Bueler.Pelt.2015}. Basal meltwater is accumulated
    locally without transportation. When the till becomes saturated, additional
    meltwater assumed to drain off instantaneously outside the glacier margins,
    i.e. it is removed from the system in an accountable way. Other parameters
    (Table~\ref{tab:params}) follow simulations of the Greenland ice sheet
    \citep{Aschwanden.etal.2013}, or benchmarks when other data is missing
    \citep{Bueler.Pelt.2015}.

@@

    The temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are
    computed as a~function of ice surface elevation, $s$, using the SRTM-based
    topography data provided with WorldClim as a~reference, $b_{\text{ref}}$:
    %
    \begin{align}
      {\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
                                     &= -\gamma [h(t, x, y)+
                                                 b(t, x, y)-b_{\text{ref}}],
    \end{align}
    %
    thus accounting for the evolution of ice thickness, ${h=s-b}$, on the one
    hand, and for differences between the basal topography of the ice flow
    model, $b$, and the WorldClim reference topography, $b_{\text{ref}}$, on
    the other hand. All simulations use an annual temperature lapse rate of
    $\gamma=6\,\unit{K\,km^{-1}}$, which is slightly above average but more
    representative of summer months when surface melt occurs
    \citep[Fig.~3]{Rolland.2003}.

@@

    \begin{figure*}[t]
      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
      \caption{%
        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
        (200\,m contours), and ice surface velocity (blue) in the Alps
        24.57\,ka before present, corresponding to the maximum modelled ice
        cover. The modelled LGM ice extent (dashed orange line) and
        geomorphological reconstruction \citep[solid red
        line,][]{Ehlers.etal.2011}, and some major transfluences across
        mountain passes (crosses) are shown.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model \citep[black
        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
        last glacial cycle (120--0\,ka), expressed in meters of sea level
        equivalent (m~s.l.e., blue curve). Shaded gray areas indicate the
        timing for Marine Oxygen Isotope Stage (MIS)~2 and MIS~4 according to
        a~global compilation of benthic \chem{\delta^{18}O} records
        \citep{Lisiecki.Raymo.2005}.}
      \label{fig:lgmvel}
    \end{figure*}

@@

    On the other hand, while atmospheric circulation models
    \citep{Strandberg.etal.2011, Ludwig.etal.2016} and palaeo-climate proxies
    \citep{Luetscher.etal.2015} both support differential precipitation changes
    north and south of the Alps, our model results show no obvious north-south
    bias as compared to the mapped LGM margins (Fig.~\ref{fig:lgmvel}a) despite
    the homogeneous temperature and precipitation anomalies applied relative to
    present. Although this may appear as a contradiction with previous,
    constant-climate modelling results \citep{Becker.etal.2016}, it follows
    from introducing a time-dependent palaeo-temperature forcing. Without
    introducing differential precipitation change north and south of the Alps,
    constant climate forcing systematically resulted in extraneous
    \citep[Fig.~3]{Becker.etal.2016} and premature
    \citep[Fig.~4]{Becker.etal.2016} glaciation on the north slope of the Alps.
    Using time-dependent palaeo-temperature forcing, progressive atmospheric
    cooling over several thousand years allow for warmer temperatures, closer to
    the thermodynamic equilibrium, within the ice and the bedrock. This results
    in thinner glaciers and limits overshoot of the equilibrium state
    (discussed further in Sect.~\ref{sec:timing}).

@@

    In the Western Alps, major transfluences occur for instance across Col de
    Montgenèvre, Col du Mont-Cenis, Simplon Pass and Brünig Pass
    (Fig.~\ref{fig:lgmvel}a). Although a transfluence across Col de Montgenèvre
    was previously questioned \citep[Fig.~2]{Cossart.etal.2012}, evidence for
    southerly ice flow across Col du Mont-Cenis has been recognized
    \citep[Fig.~3.18, p.~284]{Onde.1938, Coutterand.2010}. Similarly,
    tranfluences have been previously identified from the geomorphology across
    Simplon Pass \citep{Kelly.etal.2004} and from the distribution of glacial
    erratics across Brünig Pass \citep{Jackli.1962}. On the other hand, the
    simplified climate forcing used in this sumulation could not reproduce
    the transport of glacial erratics from southern Valais to observed
    deposition sites in the Solothurn region \citep[cf.~][]{Jouvet.etal.2017a}.

@@

    The timing of the LGM has been documented by radiocarbon and cosmogenic
    isotope dating techniques at multiple locations around the Alps (cf.
    \citealp{Ivy-Ochs.2015};
    \citealp[Fig.~5]{Wirsig.etal.2016} for reviews; and the more recent
    publications by \citealp{Monegato.etal.2017, Federici.etal.2017}). These
    data indicate that Alpine glaciers reached their maximum extent between 26
    and 20\,ka (calibrated \chem{^{14}C} and \chem{^{10}Be} ages), but also
    that terminal lobes stayed in the foreland until 22\,ka to 17\,ka,
    when they experienced rapid retreat synchronously to lowering of the ice
    surface in the mountains (\citealp{Ivy-Ochs.2015}; \citealp[Fig.~5]{Wirsig.etal.2016};
    \citealp[Fig.~3]{Monegato.etal.2017}).

@@

    In the upper Rhone Valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (and a standard deviation 197\,m,
    Fig.~\ref{fig:trimlines}b), which is significant in respect to the modelled
    maximum ice thickness in the Rhone Valley (and the Alps) of 2586\,m. This
    result depends on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here.
    However, regional model sensitivity tests show that a modelled surface
    compatible with the trimline elevations is incompatible with the documented
    overspilling flow across the Simplon Pass \citep{Becker.etal.2017}, and LGM
    palaeo-climate proxy records \citep{Cohen.etal.2017}. Unfortunately,
    validation through the bedrock uplift rate \citep[cf.][]{Kuchar.etal.2012}
    is not doable in the Alps due its lower values, active tectonics and
    uncertainties on geological properties \citep[cf.][]{Mey.etal.2016}.

@@

    For instance the Rhine Glacier (Fig.~\ref{fig:profiles}a) reaches a length
    exceeding 130\,km, roughly corresponding to the location of the last major
    Alpine reliefs and the beginning of the foothills, six times during the
    simulation, and sometimes retreats almost completely between two advances
    (Fig.~\ref{fig:profiles}b). The Rhone Glacier, fed by several high-altitude
    accumulation areas, advanced eight times onto modern Lake Geneva
    (Fig.~\ref{fig:profiles}c and d). The Dora Baltea Glacier, characterised
    by a very steep catchment, and mutliple tributaries, shows even a higher
    variability and reaches close to its maximum position six times throughout
    the simulation (Fig.~\ref{fig:profiles}e and f). The Isère
    (Fig.~\ref{fig:profiles}g and h) and Inn (Fig.~\ref{fig:profiles}i and j)
    Glaciers, with their complex system of confluences and diffluences, need
    longer time to build up and reach the foreland only two or three times
    (Fig.~\ref{fig:profiles}g--j). Finally, the Tagliamento Glacier, distant
    from the major ice-dispersal centres of the inner Alps, is absent for most
    of the modelled glacial cycle (Fig.~\ref{fig:profiles}k and l)

@@

    However, these results are limited by uncertainties on ice physics. The
    till deformation model used here does not hold for sliding over
    bedrock surfaces. On the other hand, the constant friction angle used is
    representative of wet till but weaker basal conditions may have applied
    over saturated lake sediments where they occured. In the absence of ice
    deformation measurements, a constant rheology was used for temperature ice
    containing more than 1\,\% of liquid water.

>>>>

    Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
    glacial cycle (120--0\,ka) in the Alps, using horizontal resolutions of 2
    and 1\,km. Climate forcing is derived using present-day climate data from
    WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
    offsets from multiple palaeo-climate proxies, among which only the EPICA ice
    core record yields glaciation during marine oxygen isotope stages~4
    (69--62\,ka) and~2 (34--18\,ka) spatially and temporally consistent with
    the geological reconstructions, while the other records used result in
    excessive early glacial cycle ice cover and a late Last Glacial Maximum.
    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that Alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences through the
    mountain passes. Modelled maximum ice surface is in average 861\,m
    higher than observed trimline elevations in the upper Rhone Valley, yet
    our simulation predicts little erosion at high elevation due to cold-based
    ice. Finally, despite the uniform climate forcing, differences in glacier
    catchment hypsometry produce a time-transgressive Last Glacial Maximum
    advance, with some glaciers reaching their modelled maximum extent as
    early as 27\,ka, and others as late as 21\,ka.

@@

    Although the glacial history of the European Alps has been studied for
    nearly three hundred years, uncertainties remain on
      (1) what climate evolution led to the known maximum ice limits,
      (2) to what extent ice flow was controlled by subglacial topography,
      (3) what drove the different responses of the individual lobes,
      (4) how far above the trimline was the ice surface located, and
      (5) how many advances occurred during the last glacial cycle.

    Here, we intend to explore these open questions from a new angle and
    use the Parallel Ice Sheet Model
    \citep[PISM,][]{PISM-authors.2017}, a numerical ice sheet model that
    approximates glacier sliding and deformation (Sect.~\ref{sec:model}), to
    model Alpine glacier dynamics through the last glacial cycle (120--0\,ka),
    a period for which palaeo-temperature proxies are available, albeit non
    regional. In an attempt to analyse the longstanding questions outlined
    above, we test the model sensitivity to multiple palaeo-climate forcing
    (Sect.~\ref{sec:climate}), and then explore the modelled glacier dynamics
    at high resolution for the optimal forcing (Sect.~\ref{sec:results}).
    Additional geological research will be needed to complete our knowledge.

@@

    Each simulation starts from assumed present-day ice thickness and
    equilibrium ice and bedrock temperature at 120\,ka, and runs to the
    present. Our modelling domain of 900 by 600\,km encompasses the
    entire Alpine range (Fig.~\ref{fig:inputs}). The simulations were run on
    two different grids, using horizontal resolutions of 2 and 1\,km,
    respectively.
    All parameter values are summarized in Table~\ref{tab:params}.

@@

    Ice sheet dynamics are typically modelled using a~combination of internal
    deformation and basal sliding. PISM is a~shallow ice sheet model, which
    implies that the balance of stresses is approximated based on their
    predominant components. The Shallow Shelf Approximation (SSA) is combined
    with the Shallow Ice Approximation (SIA) by adding velocity solutions of
    the two approximations \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
    In the SIA, topographic roughness is parametrized using a bed smoother
    range of 5\,km \citep{Schoof.2003}.
    Although the SSA-SIA heuristic is inaccurate in the transition zone
    where both vertical shear and longitudinal stretch are significant,
    it was shown to effectively approximate complex ice sheet flow
    dynamics over mountainous topographies similar to that of the European Alps
    \citep{Golledge.etal.2012, Ziemen.etal.2016}.

@@

    A~pseudo-plastic sliding law relates the bed-parallel shear stresses to the
    sliding velocity. The yield stress is modelled using the Mohr--Coulomb
    criterion \citep[Table~\ref{tab:params};][Eq.~18]{Tulaczyk.etal.2000}.
    using a constant basal friction angle corresponding to the average of
    available measurements \citep[p.~268]{Cuffey.Paterson.2010}.
    Effective pressure is related to the ice overburden stress and the modelled
    amount of subglacial water, using a~formula derived from laboratory
    experiments with till extracted from the base of Ice Stream B in West
    Antarctica \citep[Table~\ref{tab:params};][Eqs~23 and
    24]{Tulaczyk.etal.2000, Bueler.Pelt.2015}. Basal meltwater is accumulated
    locally without horizontal transport. When the till becomes saturated,
    additional meltwater assumed to drain instantaneously and is removed from
    the system in an accountable way. Other parameters
    (Table~\ref{tab:params}) follow simulations of the Greenland ice sheet
    \citep{Aschwanden.etal.2013}, or benchmarks when other data is missing
    \citep{Bueler.Pelt.2015}.

@@

    The temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are
    computed as a~function of ice surface elevation, $s$, using the SRTM-based
    topography data provided with WorldClim as a~reference, $b_{\text{ref}}$:
    %
    \begin{align}
      {\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
                                     &= -\gamma [h(t, x, y)+
                                                 b(t, x, y)-b_{\text{ref}}],
    \end{align}
    %
    thus accounting for the evolution of ice thickness, ${h=s-b}$, on the one
    hand, and for differences between the basal topography of the ice flow
    model, $b$, and the WorldClim reference topography, $b_{\text{ref}}$, on
    the other hand. All simulations use an annual temperature lapse rate of
    $\gamma=6\,\unit{K\,km^{-1}}$, which is slightly above annual lapse rates
    measured in the Alps but more representative of summer months when surface
    melt occurs \citep[Fig.~3]{Rolland.2003}.

@@

    \begin{figure*}[t]
      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
      \caption{%
        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
        (200\,m contours), and ice surface velocity (blue) in the Alps
        24.57\,ka before present, corresponding to the modelled maximum ice
        cover. The modelled LGM ice extent (dashed orange line) and
        geomorphological reconstruction \citep[solid red
        line,][]{Ehlers.etal.2011}, and some major transfluences across
        mountain passes (crosses) are shown.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model \citep[black
        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
        last glacial cycle (120--0\,ka), expressed in meters of sea level
        equivalent (m~s.l.e., blue curve). Shaded gray areas indicate the
        timing for Marine Oxygen Isotope Stage (MIS)~2 and MIS~4 according to
        a~global compilation of benthic \chem{\delta^{18}O} records
        \citep{Lisiecki.Raymo.2005}. The black vertical line indicates the
        modelled age of maximum ice cover at 24.57\,ka.}
      \label{fig:lgmvel}
    \end{figure*}

@@

    On the other hand, while atmospheric circulation models
    \citep{Strandberg.etal.2011, Ludwig.etal.2016} and palaeo-climate proxies
    \citep{Luetscher.etal.2015} both support differential precipitation changes
    north and south of the Alps, our model results show no obvious north-south
    bias as compared to the mapped LGM margins (Fig.~\ref{fig:lgmvel}a) despite
    the homogeneous temperature and precipitation anomalies applied relative to
    present. Although this may appear as a contradiction with previous,
    constant-climate modelling results \citep{Becker.etal.2016,
    Jouvet.etal.2017}, it follows
    from introducing a time-dependent palaeo-temperature forcing. Without
    introducing differential precipitation change north and south of the Alps,
    constant climate forcing systematically resulted in extraneous
    \citep[Fig.~3]{Becker.etal.2016} and premature
    \citep[Fig.~4]{Becker.etal.2016} glaciation on the north slope of the Alps.
    Using time-dependent palaeo-temperature forcing, progressive atmospheric
    cooling over several thousand years allow for warmer temperatures, closer to
    the thermodynamic equilibrium, within the ice and the bedrock. This results
    in thinner glaciers and limits overshoot of the equilibrium state
    (discussed further in Sect.~\ref{sec:timing}).

@@

    In the Western Alps, major transfluences occur for instance across Col de
    Montgenèvre, Col du Mont-Cenis, Simplon Pass and Brünig Pass
    (Fig.~\ref{fig:lgmvel}a). Although a transfluence across Col de Montgenèvre
    was previously questioned \citep[Fig.~2]{Cossart.etal.2012}, evidence for
    southerly ice flow across Col du Mont-Cenis has been recognized
    \citep[Fig.~3.18, p.~284]{Onde.1938, Coutterand.2010}. Similarly,
    tranfluences have been previously identified from the geomorphology across
    Simplon Pass \citep{Kelly.etal.2004} and from the distribution of glacial
    erratics across Brünig Pass \citep{Jackli.1962}. On the other hand, the
    simplified climate forcing used in this simulation could not reproduce
    the transport of glacial erratics from southern Valais to observed
    deposition sites in the Solothurn region \citep[cf.~][]{Jouvet.etal.2017a}.

@@

    The timing of the LGM has been documented by radiocarbon and cosmogenic
    isotope dating techniques at multiple locations around the Alps (cf.
    \citealp{Ivy-Ochs.2015};
    \citealp{Wirsig.etal.2016} for reviews; and the more recent
    publications by \citealp{Monegato.etal.2017, Federici.etal.2017}). These
    data indicate that Alpine glaciers reached their maximum extent between 26
    and 20\,ka (calibrated \chem{^{14}C} and \chem{^{10}Be} ages), but also
    that terminal lobes stayed in the foreland until 22\,ka to 17\,ka,
    when they experienced rapid retreat synchronously to lowering of the ice
    surface in the mountains (\citealp{Ivy-Ochs.2015};
    \citealp[Fig.~5]{Wirsig.etal.2016}; \citealp[Fig.~3]{Monegato.etal.2017}).

@@

    In the upper Rhone Valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (and a standard deviation 197\,m,
    Fig.~\ref{fig:trimlines}b), which is significant in respect to the modelled
    maximum ice thickness in the Rhone Valley (and the Alps) of 2586\,m. This
    result depends on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here.  However, regional
    model sensitivity tests show that a modelled surface compatible with the
    trimline elevations is incompatible with the documented overspilling flow
    across the Simplon Pass \citep{Becker.2017, Becker.etal.2017}, and LGM
    palaeo-climate proxy records \citep{Cohen.etal.2017}. Unfortunately,
    validation through the bedrock uplift rate \citep[cf.][]{Kuchar.etal.2012}
    is not doable in the Alps due its lower values, active tectonics and
    uncertainties on geological properties \citep[cf.][]{Mey.etal.2016}.

@@

    For instance the Rhine Glacier (Fig.~\ref{fig:profiles}a) extends beyond
    the outermost limestone reliefs six times during the
    simulation, and sometimes retreats almost completely between two advances
    (Fig.~\ref{fig:profiles}b). The Rhone Glacier, fed by several high-altitude
    accumulation areas, advances eight times onto modern Lake Geneva
    (Fig.~\ref{fig:profiles}c and d). The Dora Baltea Glacier, characterised
    by a very steep catchment, and mutliple tributaries, shows even a higher
    variability and reaches close to its maximum position six times throughout
    the simulation (Fig.~\ref{fig:profiles}e and f). The Isère
    (Fig.~\ref{fig:profiles}g and h) and Inn (Fig.~\ref{fig:profiles}i and j)
    Glaciers, with their complex system of confluences and diffluences, need
    longer time to build up and reach the foreland only two or three times
    (Fig.~\ref{fig:profiles}g--j). Finally, the Tagliamento Glacier, distant
    from the major ice-dispersal centres of the inner Alps, is absent for most
    of the modelled glacial cycle (Fig.~\ref{fig:profiles}k and l)

@@

    However, these results are limited by uncertainties on glacier physics. The
    till deformation model used here does not hold for sliding over
    bedrock surfaces. On the other hand, the constant friction angle used is
    representative of wet till but weaker basal conditions may have applied
    over saturated lake sediments where they occured. In the absence of ice
    deformation measurements, a constant rheology was used for temperature ice
    containing more than 1\,\% of liquid water.
